Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-03 talk/dls2014/paper/paper.tex
Rewrite section 3.2.  Please review...

\conferenceinfo{DLS 2014}{to be supplied}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

TM systems can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well on low numbers of CPUs.
\item We make novel use of the memory management capabilities of
  off-the-shelf CPUs, e.g. memory segmentation and 64bit address space.
\item We integrate our STM system closely with a garbage collector
  (GC) in order to lower the overhead of STM.
\item This new STM system is used to replace the GIL in one
  implementation of Python and is then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure~\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection or we start the next
transaction~\footnote{We always do a minor collection during a commit.}.

Note that we have three kinds of objects: \emph{young, overflow} and
\emph{old} objects. Young and overflow objects were created in the
same transaction; old objects always come from previous transactions.
The close integration between STM and the GC allows us to use this
information effectively: young objects do not trigger the slow path at
all, overflow objects only need to be handled by the GC part of the
write barrier once between collections, and only the old objects need
to be handled by the TM part once per transaction. This keeps the
overhead of the write barrier very low.

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing JIT\cite{cfbolz09}
\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

For the results in Figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$13\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$13\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
more iterations per second than a single thread.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in Figure~\ref{fig:performance-nojit}.

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see \cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (\emph{mandelbrot,
raytrace, richards}), and behave similar to the GIL interpreters in the
other cases. The reason again being the coarse grained locking.

PyPy using our STM system (\emph{pypy-stm-nojit}) scales in all
benchmarks to a certain degree. It scales best in the ones where
Jython scales as well and a little less in the others. The reason
for that is that in the former group there are no real, logical
conflicts -- all threads do independent calculations. In the latter
case, the threads work on a common data structure and therefore
create much more conflicts, which limits the scalability.

Looking at the average overhead from switching from GIL to STM, we see
that it is \remi{$\approx 35.5\%$}. The maximum in richards is
\remi{$63\%$}.

\emph{pypy-stm-nojit} beats \emph{pypy-nojit} already on two threads;
however, it never even beats CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. So we will now look at how well our system works when we
enable the JIT.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing execution time between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

@@

The results are presented in Figure~\ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. The slowdown factor for switching from GIL
to STM ranges around \remi{$1-2.4\times$}, and we beat GIL performance
in half of the benchmarks.

We see that generally, the group of non-locked benchmarks scales
best. The other three scale barely or not at all with the number of
threads. The reason for this is likely again the conflicts in the
latter group. Since the JIT accelerates all code but not the STM
overhead, we do more work per transaction. And this increases the
likelihood of conflicts between them and therefore limits scalability
even more than in the no-JIT benchmarks.

Overall PyPy needs the JIT in order for its performance to be
competitive. It would be interesting to see how using our STM system
in CPython would turn out, but it is a lot of work. On its own, our
system scales well, so we hope to also see that with the JIT in the
future.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
  \caption{Comparing execution time between interpreters with JIT\label{fig:performance-jit}}
\end{figure}

@@

There have been several attempts at removing the GIL using TM. We
ourselves did a few earlier attempts with more conventional STM
systems~\cite{stmupdate13}. Because of the huge overhead we
encountered, the result was mostly useless on $\le 4$ CPU cores.
Since this could be considered the main operating conditions of
dynamic languages, it was never as useful as we wanted it to be.

Other attempts using HTM~\cite{nicholas06,odaira14,fuad10} show
promising results, especially the very recent \cite{odaira14} that
examines the current generation of Intel CPUs. However, they did a
simplification by essentially disabling the GC during the
benchmarks. This may be valid for the hand-tuned allocations in their
interpreter. For us the GC is essential, and because it is a moving
GC, it unfortunately touches a lot more memory and thereby reaches the
HTM capacity much faster.  This, and the fact that HTM's capacity
limits do not allow the implementation of arbitrary atomic blocks, let
us to believe that STM is the right choice at the moment.

There are multiple attempts at removing the limits of HTM by means of
\emph{virtualising} HTM~\cite{rajwar05,chung06}. Indeed this may also
be the future direction of our research. So far, some of them depend
on additional hardware features. Or in the case of \cite{chung06},
they have the assumption that most transactions fit the HTM capacity
and that virtualisation is only a fallback. Our use case, however,
features arbitrarily long transactions, which is why this assumption
first has to be verified in our case.

From the other end, there also have been attempts at lowering the
overhead of STM~\cite{warmhoff13,spear09} especially for a low number
of CPU cores. \cite{spear09} works well for some workloads, in the
case of a dynamic language VM, we have to assume any possible
workload.  \cite{warmhoff13} uses a compile-time approach to select
from multiple optimised STMs. Their partitioned FastLane STM seems to
provide good performance already on two threads using master and
helper threads. It would be interesting to see if its design also
performs well in our use case.

Similarly to us, \cite{martin09} use memory management techniques to
provide isolation between transactions. They focus on providing strong
atomicity between transactional and non-transactional code.  We,
however, use inevitable transactions to provide strong atomicity.
Additionally, while they use page fault handlers to detect conflicts,
we use traditional write barriers. Again, there are only some
superficial similarities between our approach and theirs.

% Similar STMs:
% \begin{itemize}
% \item FastLane: \cite{warmhoff13}
% \item TML: \cite{spear09}
% \item Virtualizing HTM: \cite{rajwar05}
% \item Page-based virtualizing HyTM: \cite{chung06}: page-level conflict
%   detection, otherwise hardware extensions required; assumes most
%   transactions fit HTM capacities (not so true here); COW using page-faults;
%   they assume OS-level access to page-tables (maybe not inherent to their
%   design); eval on simulator; value-based confl detection;

%  (XTM can be
%   implemented either in the OS as part of the virtual memory manager or
%   between underlying TM systems and the OS, like virtual machines;
%   Conflicts for overflowed transactions are tracked at page granularity;
%   XTM-e allows conflict detection at cache line granu-
%   larity, even for overflowed data in virtual memory)
% \item using mmap(): Memory-Mapped Transactions
% \item mem-protected conflict detection: \cite{martin09}
% \end{itemize}

>>>>

\conferenceinfo{ICOOOLPS workshop 2014}{July 28th, 2014, Uppsala, Sweden}
\copyrightyear{2014}
%\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\doi{nnnnnnn.nnnnnnn}

@@

TM systems come in\arigo{typo?} can be broadly categorised as hardware based (HTM),
software based (STM), or hybrid systems (HyTM). HTM systems are limited
by hardware constraints~\cite{odaira14,fuad10}, while STM systems have
a lot of overhead~\cite{cascaval08,drago11}. In \cite{wayforward14},
we argue that STM is still the best way forward, especially since it
supports large atomic blocks as a new way for synchronising multiple
threads. There have been several attempts at lowering the overhead
of STM~\cite{warmhoff13,spear09} -- sometimes at the cost of scalability.
In this paper, we describe how we manage to lower the overhead of our
STM system so that it can be seen as a viable replacement for the GIL.

Our contributions include:
\begin{itemize}[noitemsep]
\item We introduce a new software transactional memory (STM) system
  that performs well even on low numbers of CPUs. It uses a novel
  combination of long-existing CPU features and garbage collector (GC)
  integration in order to keep the overhead of STM very low.
\item This new STM system is used to replace the GIL in one implementation
  of Python and is then evaluated extensively.
\item We introduce atomic blocks to the Python language to provide a
  backwards compatible, composable synchronisation mechanism for
  threads.
\end{itemize}

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
figure~\ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

In all cases, we remove the \lstinline!WRITE_BARRIER!  flag from the
object before we return. Thus, we never trigger the slow path again
before we do the next minor collection or we start the next
transaction (we always do a minor collection during a commit).

@@

Here, we will not go into detail about the integration of our STM
system with PyPy's JIT. In fact, we will disable it for all benchmarks
except those in section~\ref{sec:performance-bench}. We would like to
regard it as a simple performance enhancement, but that is not what
happens in reality. First, since the JIT is a tracing
JIT\remi{explain?}  running in multiple threads, it may compile
different things in each run because of the non-deterministic
thread-scheduling of the operating system (OS). Second, it is able to
remove some allocations in some cases. Because compilation is already
non-deterministic, so is this allocation-removal. And third, we did
not have enough time to optimise integration with STM so that the JIT
exposes the overhead of STM more by speeding up all the rest.

@@

In figure~\ref{fig:richards_mem} we look at the memory usage of one of
our benchmarks called Richards\footnote{OS kernel simulation
benchmark}. The \emph{GC managed memory} counts all memory used in the
old object space including the memory required for private pages. The
sharp drops in memory usage come from major collections that free old
objects and re-share pages. The average memory usage stays around
29~MiB and we see that in this benchmark we have around 1 major
collection every second.

@@

For the results in figure~\ref{fig:scaling}, we averaged over 5 runs
and normalised the average runtimes to the time it took on a single
thread. From this we see that there is additional overhead introduced
by each thread (\remi{$13\%$} for all 4 threads together). While this
is not ideal, we think that \remi{$13\%$} are acceptable on four
threads. In terms of throughput, 4 threads have \remi{$3.53\times$}
more iterations per second than a single thread.

@@

\paragraph{Non-JIT benchmarks:} First we run our benchmarks on four
different interpreters: Jython (fine-grained locking), CPython (GIL),
and PyPy with STM and with the GIL (both without the JIT). The results
are shown in figure~\ref{fig:performance-nojit}.

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see \cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (mandelbrot,
raytrace, richards), and behave similar to the GIL interpreters in the
other cases.

PyPy using our STM system (pypy-stm-nojit) scales in all benchmarks to
a certain degree. We see that the average overhead from switching from
GIL to STM is \remi{$35.5\%$}, the maximum in richards is
\remi{$63\%$}. pypy-stm-nojit beats pypy-nojit already on two threads;
however, it never even beats CPython, the reference implementation of
Python. This means that without the JIT, our performance is not
competitive. We now look at how well our system works when we enable
the JIT.

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance_nojit.pdf}
  \caption{Comparing runtime between interpreters without a JIT\label{fig:performance-nojit}}
\end{figure}

@@

The results are presented in figure~\ref{fig:performance-nojit}. We
see that the performance is much less stable. There is certainly more
work required in this area. In general, we see that the group of
non-locked benchmarks certainly scales best. The other three scale
barely or not at all with the number of threads. The slowdown factor
from GIL to STM ranges around \remi{$1-2.4\times$} and we beat GIL
performance in half of the benchmarks.

\remi{Reason for bad scaling: acceleration of code that produces
conflicts $-->$ more iterations $-->$ more conflicts. The overhead
doesn't get accelerated by the JIT.}

\begin{figure}[h]
  \centering
  \includegraphics[width=1\columnwidth]{plots/performance.pdf}
  \caption{Comparing runtime between interpreters with JIT\label{fig:performance-jit}}
\end{figure}

Overall PyPy needs the JIT in order for its performance to be
competitive.  It would be interesting to see how using our STM system
in CPython would perform, but it is a lot of work. On its own, our
system scales well so we hope to also see that with the JIT in the
future.

@@

Eliminate GIL:
\begin{itemize}
\item Previous attempts with HTM: \cite{nicholas06,odaira14,fuad10}
\item Previous attempts with STM: \cite{stmupdate13}
\end{itemize}

Similar STMs:
\begin{itemize}
\item FastLane: \cite{warmhoff13}
\item TML: \cite{spear09}
\item Virtualizing HTM: \cite{rajwar05}
\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
  implemented either in the OS as part of the virtual memory manager or
  between underlying TM systems and the OS, like virtual machines;
  Conflicts for overflowed transactions are tracked at page granularity;
  XTM-e allows conflict detection at cache line granu-
  larity, even for overflowed data in virtual memory)
\item using mmap(): Memory-Mapped Transactions
\item mem-protected conflict detection: \cite{martin09}
\end{itemize}
